---
title: "Matrices"
---

## Objectives

This chapter explores fundamental matrix concepts, including operations, applications in summarizing data, and their role in multiple linear regression.

- Review of matrix operations  
- Special cases of matrix multiplication  
- Summarizing multiple variables (multivariate normal distribution)  
- Revisiting multiple linear regression (MLR) 


## Review of Matrix Operations

### Symmetric Matrices  
- A symmetric matrix has the same number of rows and columns (i.e a square matrix).  

### Vectors  
- A vector is a matrix with a single column.  
- Variables in vector format: $X = (X_1, X_2, X_3, X_4)$

### Transposing Matrices  
- The transpose of a matrix swaps its rows and columns.  
- If $A$ is a matrix, then its transpose is denoted as $A'$ or $A^T$.  
- The first column becomes the first row in the transpose.  

### Matrix Addition and Subtraction  
- Matrices must have the same dimensions for addition or subtraction.  
- Operations are performed elementwise. 
 
### Matrix Multiplication  

- Not all matrices can be multiplied—matrix multiplication is only defined if the number of columns in the first matrix matches the number of rows in the second matrix.  
- If $A$ is an $m \times n$ matrix and $B$ is an $n \times p$ matrix, then the product $AB$ is an $m \times p$ matrix.  
- Each element in $AB$ is obtained by computing the dot product of a row from $A$ and a column from $B$.  

#### What is a Dot Product?  
The dot product of two vectors is the sum of the element-wise multiplications of their corresponding entries.  

For matrix multiplication, to compute the element at row $i$, column $j$ of $AB$, we take:  
1. Row $i$ from $A$  
2. Column $j$ from $B$  
3. Multiply corresponding elements and sum them:  
   $$
   AB_{i,j} = A_{i,1}B_{1,j} + A_{i,2}B_{2,j} + \dots + A_{i,n}B_{n,j}
   $$  
 
#### Step-by-Step Example  

Let:  
$$
A =
\begin{bmatrix}
1 & 2 & 3 \\
4 & 5 & 6
\end{bmatrix}
$$
($2 \times 3$ matrix)

$$
B =
\begin{bmatrix}
7 & 8 \\
9 & 10 \\
11 & 12
\end{bmatrix}
$$
($3 \times 2$ matrix)

Since $A$ has **3 columns** and $B$ has **3 rows**, multiplication is valid, and the resulting matrix $AB$ has dimension $2 \times 2$.  

To compute the element in **row 1, column 1** of $AB$:  

$$
AB_{1,1} = (1 \times 7) + (2 \times 9) + (3 \times 11) = 7 + 18 + 33 = 58
$$

To compute the element in **row 1, column 2**:  

$$
AB_{1,2} = (1 \times 8) + (2 \times 10) + (3 \times 12) = 8 + 20 + 36 = 64
$$

To compute the element in **row 2, column 1**:  

$$
AB_{2,1} = (4 \times 7) + (5 \times 9) + (6 \times 11) = 28 + 45 + 66 = 139
$$

To compute the element in **row 2, column 2**:  

$$
AB_{2,2} = (4 \times 8) + (5 \times 10) + (6 \times 12) = 32 + 50 + 72 = 154
$$

Thus, the final matrix product is:  

$$
AB =
\begin{bmatrix}
58 & 64 \\
139 & 154
\end{bmatrix}
$$  

### Identity Matrices  
- An identity matrix is always symmetric, with:  
  - All diagonal elements of **1**  
  - All off-diagonal elements of **0** 
- The identity matrix behaves like **1** in scalar multiplication:  
  - If $A$ is an $r \times c$ matrix and $I$ is a $c \times c$ identity matrix, then $AI = A$.  
  - If $I$ is an $r \times r$ identity matrix, then $IA = A$.  

### Matrix Inverses  
- If a matrix is square and meets certain conditions, an inverse matrix exists.  
- The inverse of $A$ is denoted as $A^{-1}$.  
- The property: $ AA^{-1} = A^{-1}A = I $  
- Inverse operations are matrix equivalents of division.  
  
### Tips for Reading Matrix Formulas  
- Always check the dimension of the final result when interpreting matrix expressions.  

---

## Special Cases of Matrix Multiplication

Let $C_{n \times 1}$ be a column vector of chosen numbers, and let $Y_{n \times 1}$ be a column vector representing a sample of data. The computation $C' Y$ (the transpose of $C$ multiplied by $Y$) allows us to compute:

### **Averages**
If all elements of $C$ are set to $\frac{1}{n}$, then:
$$
C' Y = \frac{1}{n} \sum Y_i = \bar{Y}
$$
This computes the **sample mean** of $Y$.

### **Weighted Averages**
If $C$ contains weights that sum to 1 and are all positive, then:
$$
C' Y = \sum w_i Y_i
$$
This computes a **weighted average** of $Y$.

### **Summation**
If all elements of $C$ are set to 1, then:
$$
C' Y = \sum Y_i
$$
This computes the **sum** of all values in $Y$.

### **Extending to Matrices**
If $C$ is not a vector but a matrix, these computations are performed simultaneously for multiple variables using **matrix multiplication**:
$$
C' Y =
\begin{bmatrix}
\frac{1}{n} & \frac{1}{n} & \dots & \frac{1}{n} \\
1 & 1 & \dots & 1
\end{bmatrix}
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_n
\end{bmatrix}
$$
This results in a matrix where one row stores the means of $Y$ and another stores the sums.

### **Sums of Squares**
Using matrix multiplication:
$$
Y' Y = \sum Y_i^2
$$
This computes the **sum of squares** of $Y$, which is often used in variance and regression calculations.


## Summarizing Multiple Variables

- Summarizing multiple numeric variables  
  - One-variable summaries  
    - Mean and standard deviation (for normally distributed data)  
    - Five-number summary (Min, 1st quartile, Median, 3rd quartile, Max)  
  - Multiple variables  
    - Mean/standard deviation for each variable  
    - Variables might be correlated as well  
    - Matrices can help us "bookkeep" the information  
    
- Parameters (Two-Variable Case)  
  - Variable 1 (X~1~)  
    - Mean $\mu_1$  
    - Standard deviation $\sigma_1$  
  - Variable 2 (X~2~)  
    - Mean $\mu_2$  
    - Standard deviation $\sigma_2$   
  - Additional parameter needed to tell us how the variables depend on each other (covary with each other)  
    - Covariance $\sigma$~12~  
    - Directly related to correlation  
    
- Matrix Representation  
  - X~1~ and X~2~ follow a multivariate distribution with a mean (vector) $\mu$ and a covariance matrix $\Sigma$.  
  
- Covariances to Correlation  
  - Correlations can be computed directly from the covariance matrix.  
    $$COR(X_1,X_2) = COV(X_1,X_2) / \sigma_1\sigma_2$$  
  - If the covariance is zero, then the variables aren't (linearly) correlated.  
  
- Covariances are "nonstandardized" correlations.  
  - Keep in mind the variables and hence the covariance has a unit of measure.  
  - The correlation depends on the covariance and the standard deviations. The raw covariance value doesn't indicate correlation directly because of scale.  
- Correlations are "standardized" covariances.  
  - They are unitless.  
  - They are between -1 < cor < 1.
  
- Multivariate Normal Distribution  
  - Multiple variables  
    - Has a mean vector and covariance matrix  
    - Has a joint distribution that categorizes the likelihood of specific combinations of observations  
- Some data sets will behave in a way that is consistent with the multivariate normal distribution (MVN).  

- MVN Distribution (Theoretical) Properties  
  - Each individual variable will behave like a normally distributed variable.  
  - The relationship between the variables will be linear.  
  - Data points will be denser around the mean vector (point).  
  
- Estimating Parameters from Data  
  - Estimates of the mean vector and covariance matrix can be obtained from a sample.  
  - Notation  
    - $\bar{X}$ is the sample mean vector.  
    - $S$ or $\hat{\Sigma}$ is the sample covariance matrix.  

- Assessing MVN  
  - Can always investigate one variable at a time, as they should be normally distributed.  
  - Check scatterplots (matrices) to make sure they are pairwise linear.  
  - QQ style plots exist.  
  - Hypothesis tests for multivariate normality also exist (null is multivariate, reject is not).  
    - Royston's test  
    - Mardia's test. 
    - There is no optimal test.  
    
- MVN Applications  
  - Discriminant analysis (classification tools)  
  - Used in regression tools  
    - Re-express assumptions in MLR  
    - Used in computations of MLR for testing regression coefficients.  
    - Repeated measures and time series  
 
- If you are interested to know if a dataset is MVN, start with scatterplot matrices.  
  - The individual curves should be bell shaped.  
  - The scatterplots should be point clouds, elliptical  or circular, dense in the middle, more linearly related for correlated variables, but no exotic trends.  
- You can also do a QQ plot (mqqnorm(dataset)), which should be highly linear on the left side with some deviations on the right side.


## MLR Revisited

How can linear regression use matrix operations and the assumption of multivariate normality?

- Simple Linear Regression  
  - The model is often loosely expressed as: $y = \beta_0 + \beta_1x + \epsilon$  
  - When building a model, the assumption is that this relationship exists for *each observation* in the data set, so technically the model is specified by n equations.  
    $y_1 = \beta_0 + \beta_1x_1 + \epsilon_1$  
                      ...  
    $y_n = \beta_0 + \beta_1x_n + \epsilon_n$  
    - The errors vary and are unique to each observation, but they are assumed to be independent, normally distributed, and have constant variance.  
    - The equations can be written in matrix notation.  
  - The errors can be treated as n separate variables.  
  - Assume that the errors are MVN.

- Big Picture   
  - For any MLR problem, to perform tests on each coefficient, need to make two computations.  
    1. Estimate the regression coefficients with: $\hat{\beta} = (X^T X)^{-1} X^T Y$  
    2. Get the SEs for the coefficients with: $\sigma^2 (X^T X)^{-1}$, and take the square root of the diagonals.  
    3. Calculate the t-statistic, p-values, and confidence with the estimates and SEs.  

- The Matrix Advantage  
  - The formula for regression coefficients is always $\hat{\beta} = (X^T X)^{-1} X^T Y$  
  - It doesn't depend on number of predictors or sample size.  


Code examples:
```{r matrices, eval = FALSE}

M = matrix(c(0,1,2,3)2,2) #2x2(rxc) matrix, adds values by column
N = matrix(c(4,3,2,1)2,2)
t(M) # transposes the matrix
M * N # elementwise multiplication
M %*% N # matrix multiplication
Minv = solve(M) # one of my functions that will find the inverse matrix
M %*% Minv # confirm it is the inverse function, this should return the identity matrix c(1,0,0,1)

```

```{r verifyingRegCoef}

# Calculate betas with matrix multiplication
x = c(1,2,3,4,5)

# create response vector
y = c(6.5, 10.8, 14, 21.2, 26.8)

# create design matrix
bigX = cbind(rep(1,5),x) # intercept all 1s, x= b1 slope
bigX

beta_hat = solve(t(bigX) %*% bigX) %*% t(bigX) %*% y
beta_hat

# compare to lm
fit = lm(y~x)
coef(summary(fit))[,1]
summary(fit)


# Calculate SE of the estimates with matrix multiplication

#get sigma^2 = MSR = (residual standard error)^2 from the model summary
cov_beta = (1.261^2) * solve(t(bigX) %*% bigX)
cov_beta

sqrt(diag(cov_beta))

# compare to lm SEs
coef(summary(fit))[,2]

```
